Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-14 SCP2008/paper/metamonads.tex
mostly done

The type |wmatrix| is for convenience, a bundling of several pieces of
data needed by both |OUTPUT| and |PIVOT| aspects:
\begin{code}
type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) rep; 
                   numcol: ('a,int) rep}
\end{code}

@@

\subsection{Main generation}\label{sec:main-generation}
We combine all user-selectable aspects in a record, which serves as a ``keyword
argument'' list to the |GenGE| functor below.
\begin{code}
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
\end{code}
%
We first instantiate the container |GVC_I| for a vector over an
integer domain.  |GenGE| is then one functor of many algorithm
generators from a linear algebra library; since all our linear algebra codes
are parameterized by the container and the domain included in it,
We pass the container to the main functor |GenLA|.
We open the resulting |G_GVC_I| module
to make all its components easily available. One of those components
is the GE generator, |GE| (which we open as well). Besides |GE|, the
module |G_GVC_I| includes the generator for GE-based solvers.  The main
generator functor is |GenGE|, shown later.  We instantiate the
generator by passing to the functor the ``record'' of various aspects; the
order is irrelevant, but all aspects such as |Det| must be
specified. In the code above, we request GE generation with full
pivot, fraction-free update, operate on non-augmented matrix and
return the |U| factor, determinant and the rank. This choice
corresponds to our sample algorithm of Fig.\ref{fig:GE-algorithm}.
We run the monad
passing the initial state |[]| and obtain the code, which we can see
by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
or with offshoring\cite{offshoring}. The code for |resIV5|
(App.~\ref{app:code-GenIV5}) shows full
pivoting, determinant and rank tracking. The code for all these
aspects is fully inlined; no extra functions are invoked and no tests
other than those needed by the LU algorithm itself are performed. The
resulting function returns a triple |int array * int * int| of the
U-factor, determinant and the rank. It is instructive to compare the
generated code with the corresponding code in
Fig.~\ref{fig:GE-algorithm}, the hand-written implementation of the
textbook pseudo-code; the only difference is the naming of
variables and the inlining of pivoting and swapping functions.

@@

The LU generator functor (called |GenGE| below) itself is
parameterized by the domain, container, pivoting policy (full, row,
nonzero, no pivoting), update policy (with either `fraction-less' or
full division), and the output specification. Some of the argument
modules such as |PIVOT| are functors themselves (parameterized by the
domain, the container, and the determinant functor).  We rely on
module subtyping: For example, |F.Output| of the type |OUTPUT| is a
functor requiring an argument structure of the signature
|OUTPUTDEP|. The fact that the signature |FEATURES| contains all the
fields of |OUTPUTDEP| and then some lets us pass |F| (of the type
|FEATURES|) as an argument to instantiate |F.Output|.

@@

There are more preflight checks for various ``semantic'' constraints, shown
in the following structure of the |UPDATE| signature:
\begin{code}
module DivisionUpdate(Det:DETERMINANT) =
  struct
  open C.Dom = struct 
  let _ = assert (C.Dom.kind = Domains_sig.Domain_is_Field)
end
\end{code}
%
This structure implements an update policy relying on
unrestricted |Dom.div|. Many domains provide |Dom.div|, for example,
the integer domain. The latter however assumes that division is applied
only if the dividend is the exact multiple of the divisor. Thus if we
specified |module Update = DivisionUpdate| when instantiating |GenIV5|
above, we would have received an error because |IntegerDomainL| is not
a field. That run-time error occurs before any code is produced (i.e.,
before |resIV5| is computed).

@@

There is an interesting relation with aspect-oriented code
\cite{kiczales97aspectoriented}: in AspectJ, 
aspects are (comparatively) lightly typed, and are post-facto extensions of
potential program traces, specified in a particular language; these tend
to be created to follow the operational behaviour of existing code, but
is not restricted to such a setting.
In our work, aspects are weaved together ``from scratch'' to
make up a piece of code.  One can understand previous work to be
more akin to dynamically typed, and dynamically specified aspect weaving, while
we have started investigating statically typed and statically specified
aspect weaving.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. The implementation of the monadic notation, |perform|,
was joint work with Lydia van Dijk.
Many helpful suggestions by the anonymous reviewers are gratefully
acknowledged.

>>>>

The type |wmatrix| denotes the |(0,0)-(numrow-1,numcol-1)|
rectangular block of |matrix|:
\begin{code}
type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) rep; 
                   numcol: ('a,int) rep}
\end{code}
The type is used in |INPUT|, |OUTPUT| and |PIVOT| aspects to refer to
the non-augmented part of the matrix.

@@

\subsection{Main generation}\label{sec:main-generation}

This section presents the core GE algorithm |GenGE|, after all aspects
have been factored out. We also describe how to instantiate the core
algorithm with sample aspects, to obtain particular GE procedures such
as our running example, Fig.~\ref{fig:GE-algorithm}. The instantiation
process is multi-step, mainly for the sake of speed of OCaml
compilation. Currently OCaml is quite slow compiling source code with
many complex modules. The benefit of the multi-step process, simple
module signatures, may help make instantiation process more
comprehensible.

The core GE algorithm and all the aspects are part of one large
functor, |Ge.LAMake|, parameterized by the interpretation aspect,
\S\ref{sec:lifteddomains}. If we select as an interpretation
generating code in the form of MetaOCaml code values, we write
\begin{code}
module GEF = Ge.LAMake(Code)
open GEF
open Domains_code
\end{code}
Here, |Code| is the interpretation with |('a,'b) rep| being 
|('a,b) code|; the module |Domains_code| contains various instances of |DOMAINL|
(such as \linebreak|FloatDomainL|, |IntegerDomainL|) for that particular
choice of |('a,'b) rep|. Opening |GEF| makes available various
container functors and the functor |GenLA|. Combining
the container functors with domain structures gives containers
with particular elements and particular representation. For example, in
\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
\end{code}
we define |GVC_I| to represent a matrix with integer elements arranged
in a flat vector in the row-major order. Instantiating the functor |GenLA|
with this container makes available all algorithmic and
tracking aspects of GE (such as |AbstractDet|, |NoDet|, |FullPivot|,
etc) as well as the module |GE| with the core GE functor |GenGE|
(besides |GE|, |GenLA| contains modules for GE-based solvers). All
these components already incorporate the choices for the
container, domain and interpretation aspects we have made earlier. We
may now select particular features of the desired GE algorithm and
instantiate and run |GenGE|.

We combine all user-selectable aspects in a record, which serves as a
``keyword argument'' list to the |GenGE| functor, shown later.
\begin{code}
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

We instantiate |GenGE| by passing to the functor the ``record'' of
various aspects; the order is irrelevant, but all aspects such as
|Det| must be specified. In the following code, we request GE generation
with full pivot, fraction-free update, operate on non-augmented matrix
and return the |U| factor, determinant and the rank. This choice
corresponds to our sample algorithm of Fig.~\ref{fig:GE-algorithm}.
\begin{code}
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
\end{code}
We run the monad passing the initial state |[]| and obtain the code,
which we can see by printing |resIV5|. The code can then be
``compiled'' as |!. resIV5| or with offshoring \cite{offshoring}. The
code for |resIV5| (App.~\ref{app:code-GenIV5}) shows full pivoting,
determinant and rank tracking. The code for all these aspects is fully
inlined; no extra functions are invoked and no tests other than those
needed by the GE algorithm itself are performed. The resulting
function returns a triple |int array * int * int| of the U-factor,
determinant and the rank. It is instructive to compare the generated
code with the corresponding code in Fig.~\ref{fig:GE-algorithm}, the
hand-written implementation of the textbook pseudo-code; the only
difference is the naming of variables and the inlining of pivoting and
swapping functions.

@@

The core GE algorithm |GenGE|, as part of |Ge.LAMake| and |GenLA|, is
already parameterized by the domain, container and interpretation.
The algorithm is further parameterized by |FEATURES|, i.e., pivoting
policy (full, row, nonzero, no pivoting), update policy (with either
`fraction-less' or full division), determinant, permutation matrix,
input and output specifications. Some of the argument modules such as
|PIVOT| are functors themselves (parameterized by the domain, the
container, and the determinant).  We rely on module subtyping: For
example, |F.Output| of the type |OUTPUT| is a functor requiring an
argument of the signature |OUTPUTDEP|. The fact that the signature
|FEATURES| contains all the fields of |OUTPUTDEP| and then some lets
us pass |F| (of the type |FEATURES|) as an argument to instantiate
|F.Output|.

@@

There are more preflight checks for various ``semantic'' constraints, shown
in the following structure of the |UPDATE| signature:
\begin{code}
module DivisionUpdate(Det:DETERMINANT) = struct
  open C.Dom
  type 'a in_val = 'a vc
  let update bic brc brk bik setter _ = perform
      y <-- ret (bik -^ ((divL bic brc) *^ brk));
      ret (setter (applyMaybe normalizerL y))
  let update_det v = Det.acc_magn v
  let upd_kind = DivisionBased
  let _ = assert (C.Dom.kind = Domains_sig.Domain_is_Field)
end
\end{code}
%
This structure implements an update policy relying on
unrestricted |Dom.divL|. Many domains provide |divL|, for example,
the integer domain. The latter however assumes that division is applied
only if the dividend is the exact multiple of the divisor. Thus if we
specified |module Update = DivisionUpdate| when instantiating |GenIV5|
above, we would have received an error because |IntegerDomainL| is not
a field. That run-time error occurs before any code is produced (i.e.,
before |resIV5| is computed).

@@

There is an interesting relation with aspect-oriented code
\cite{kiczales97aspectoriented}: in AspectJ, 
aspects are (comparatively) lightly typed, and are post-facto extensions of
potential program traces, specified in a particular language; these tend
to be created to follow the operational behaviour of existing code, but
are not restricted to such a setting.
In our work, aspects are weaved together ``from scratch'' to
make up a piece of code.  One can understand previous work to be
more akin to dynamically typed, and dynamically specified aspect weaving, while
we have started investigating statically typed and statically specified
aspect weaving.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. The implementation of the monadic notation, |perform|,
was joint work with Lydia van Dijk.  We gratefully acknowledge
numerous, helpful suggestions by the anonymous reviewers.
